Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-11 rnni_convexity.tex
everything up to Cluster Theorem looks okay now

\newcommand{\np}{\mathcal{NP}}
\newcommand{\parent}{\mathrm{parent}}
\newcommand{\mrca}{\mathrm{mrca}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\nni}{\mathrm{NNI}}
\newcommand{\rnni}{\mathrm{RNNI}}
\newcommand{\rnniu}{\mathrm{RNNIu}}
\newcommand{\tbr}{\mathrm{TBR}}
\newcommand{\spr}{\mathrm{SPR}}
\newcommand{\csort}{\textsc{Caterpillar Sort}}
\newcommand{\findpath}{\textsc{FindPath}}
\newcommand{\mdtree}{\textsc{MDTree}}
\newcommand{\rad}{\textsc{R}}

@@

\begin{algorithm}[H]
\caption{$\mdtree(T)$}
\label{alg:max_dist_tree}
\begin{algorithmic}[1]
\STATE Construct list $L=[l_1, \ldots, l_n]$ of all taxa ordered with respect to the ranks of their parents in $T$
\label{alg:mdtree_lineL}
\STATE Build tree $R$ with two taxa $l_1, l_2$ that are children of the root
\FOR {$i = n, \ldots, 3$}
\STATE Extend $R$ by attaching $l_i$ to a newly created node of rank $1$ on an arbitrarily chosen edge
\label{alg:mdtree_lineAddTaxon}
\ENDFOR
\RETURN $R$
\end{algorithmic}
\end{algorithm}

@@

The study of shortest paths in the $\rnni$ graph in this section is motivated by our aim to understand the geometry of $\rnni$ as a metric space.
We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
Those algorithms will, for example, enable us to prove that every pair of caterpillar trees $T$ and $R$ is connected by a caterpillar path of length $d(T,R)$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} $\rad(\rnni) = \min\limits_T \max\limits_R d(T,R)$.
Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
We will give a counterexample to the Split Theorem in $\rnni$ and consider a variant of the Weak Split Theorem, which states that every shared split of taxa is maintained along a shortest path.
We will call this statement the Clade Theorem, as it is slightly more natural to work with ranked trees in terms clades rather than splits (see Conjecture~\ref{conjecture:cluster_theorem}).

@@

\proof
Note that it is suffice to prove that $d_c(T, R) = d(T, R)$ for an arbitrary pair of caterpillar trees $T$ and $R$.
Lemma~\ref{lemma:caterpillar_dist=diameter} implies that if $T$ and $R$ are at the maximum possible distance, that is, $d_c(T, R) = \frac{(n-1)(n-2)}{2}$.
Denote the latter number by $D$.

@@

\proof
Since $d_c([1,2,3,\ldots,n], [n-1,n,n-2,n-3, \ldots, 1]) = \frac{(n-1)(n-2)}{2}$, the claim follows directly from from Theorem~\ref{thm:caterpillar_convex} and Lemma~\ref{lemma:diameter_bound}.
\endproof

@@

\begin{proof}
Recall that with Theorem~\ref{thm:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
Within this proof we assume, without loss of generality, that it is $T = [1,2,3,\ldots,n]$.
We prove the lemma by induction on the number of taxa $n$.

The lemma is true for the induction basis on $n=3$ taxa since all three trees have distance one, which equals diameter and radius in this case.

Let us now assume for the induction step that $T$ is a tree on $n+1$ taxa.
Taxon $n+1$ is the last one in the list $L$ computed in $\mdtree$ as its parent in $T$ is the root.
It follows that this taxon $n+1$ is the last one added to $R$ in the algorithm, so its parent has rank one in $R$.
With Lemma~\ref{lemma:distance_delete_taxon} we can follow that for $T$ and $R$ restricted to taxa $\{1, \ldots, n\}$ it holds $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
It is not hard to see that $R|_n$ can be received from $T|_n$ by applying $\mdtree$:
$R|_n$ can be constructed from input tree $T|_n$ in the same way as $R$ can be constructed from $T$, since the list $L$ computed for $T|_n$ in $\mdtree$ is the same as the one for $T$, but without the last element.
So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
\end{proof}

In Section~\ref{alg:max_dist_tree} we already discussed that $\mdtree$ is a non-deterministic algorithm.
In fact, the output tree of $\mdtree$ can have any tree topology, since in each step of the algorithm the next taxon can be added on any edge incident to a leaf in the tree.
Therefore, if the input tree is a caterpillar tree, $\mdtree$ can compute a tree with any tree topology that has maximum distance from the input tree.
And as permuting labels of trees does not change the distance between them, there exists for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
By this we obtain Theorem~\ref{thm:radius}.

\begin{theorem}
The radius of the $\rnni$ graph is $\rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
\label{thm:radius}
\end{theorem}

>>>>

\newcommand{\np}{\mathcal{NP}}
\newcommand{\parent}{\mathrm{parent}}
\newcommand{\mrca}{\mathrm{mrca}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\nni}{\mathrm{NNI}}
\newcommand{\rnni}{\mathrm{RNNI}}
\newcommand{\rnniu}{\mathrm{RNNIu}}
\newcommand{\tbr}{\mathrm{TBR}}
\newcommand{\spr}{\mathrm{SPR}}
\newcommand{\csort}{\textsc{Caterpillar Sort}}
\newcommand{\findpath}{\textsc{FindPath}}
\newcommand{\mdtree}{\textsc{MDTree}}

@@

\begin{algorithm}[H]
\caption{$\mdtree(T)$}
\label{alg:max_dist_tree}
\begin{algorithmic}[1]
\STATE Construct list $L=[l_1, \ldots, l_n]$ of all taxa ordered with respect to the ranks of their parents in $T$
\label{alg:mdtree_lineL}
\STATE Build tree $R$ with two taxa $l_1, l_2$ that are children of the root
\FOR {$i = 3, \ldots, n$}
\STATE Extend $R$ by attaching $l_i$ to a newly created node of rank $1$ on an arbitrarily chosen edge
\label{alg:mdtree_lineAddTaxon}
\ENDFOR
\RETURN $R$
\end{algorithmic}
\end{algorithm}

@@

The study of shortest paths in the $\rnni$ graph in this section is motivated by our aim to understand the geometry of $\rnni$ as a metric space.
We will employ the algorithms developed in the previous section to aid our analysis of shortest paths.
Those algorithms will, for example, enable us to prove that every pair of caterpillar trees $T$ and $R$ is connected by a caterpillar path of length $d(T,R)$ (Theorem~\ref{thm:caterpillar_convex} in Section~\ref{section:caterpillar_convex}).
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
In Section~\ref{section:diameter} we will investigate the diameter and radius of the $\rnni$ space.
Interestingly, the \emph{diameter} of $\rnni$, that is the maximum distance $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$ between any two trees in the graph, equals its \emph{radius} defined as $\min\limits_T \max\limits_R d(T,R)$.
Next, we will consider the so-called Split Theorem \autocite{Gavryushkin2018-ol}, which states that every shared split of taxa is maintained along every shortest path.
We will give a counterexample to the Split Theorem in $\rnni$ and consider a variant of the Weak Split Theorem, which states that every shared split of taxa is maintained along a shortest path.
We will call this statement the Clade Theorem, as it is slightly more natural to work with ranked trees in terms clades rather than splits (see Conjecture~\ref{conjecture:cluster_theorem}).

@@

\proof
Note that it is suffice to prove that $d_c(T, R) = d(T, R)$ for an arbitrary pair of caterpillar trees $T$ and $R$.
Lemma~\ref{lemma:caterpillar_dist=diameter} implies that if $T$ and $R$ are at the maximal possible distance, that is, $d_c(T, R) = \frac{(n-1)(n-2)}{2}$.
Denote the latter number by $D$.

@@

\proof
Since $d_c([1\ldots,n], [n-1,n,n-2,n-3, \ldots, 1]) = \frac{(n-1)(n-2)}{2}$, the claim follows directly from from Theorem~\ref{thm:caterpillar_convex} and Lemma~\ref{lemma:diameter_bound}.
\endproof

@@

\begin{proof}
We prove the lemma by induction on the number of taxa $n$.
The induction basis for $n = 3$ is trivial.

Assume now that $T = [1, \ldots, n+1]$.
In this case, the parent of taxon $n+1$ has rank $n$ in $T$ and rank $1$ in $R$.
Lemma~\ref{lemma:distance_delete_taxon} then implies that
\begin{equation}
d(T\big|_n,R\big|_n) \leq d(T,R) - (n-1).
\label{eqn:max_dist_cat_proof}
\end{equation}
Note that $R\big|_n = \mdtree (T\big|_n)$ for an appropriate choice of attachment points in the execution of $\mdtree$.
By induction hypothesis and Theorem~\ref{thm:diameter}, $d(T\big|_n,R\big|_n) = \frac{(n-1)(n-2)}{2}$.
So inequality~(\ref{eqn:max_dist_cat_proof}) becomes $d(T,R) \geq \frac{(n-1)(n-2)}{2} + n - 1 = \frac{n(n-1)}{2}$, which with Theorem~\ref{thm:diameter} gives the desired equality.
\end{proof}

The non-determinism of $\mdtree$ can be exploited to investigate what trees are at the maximal possible from a given caterpillar tree.
In fact, those trees can have an arbitrary (non-ranked) topology!
This is because in each step of the algorithm the next taxon can be added on any edge incident to a leaf in the running tree.
Therefore, if the input tree is a caterpillar tree, $\mdtree$ can be used to find a tree of pre-defined topology that is at the maximal possible distance from the input tree.
Because the distance between trees is invariant under permutations of taxa labels, we conclude that for every tree $R$ on $n$ taxa there exists a caterpillar tree $T$ such that $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
This proves the following Theorem~\ref{thm:radius}.

\begin{theorem}
The radius and diameter of the $\rnni$ space coincide and are equal to $\frac{(n-1)(n-2)}{2}$.
\label{thm:radius}
\end{theorem}
